---
title: "Alzheimer's_disease_and_Down_syndrome"
author: "Yuliia Babkova"
date: "2024-09-01"
output: html_document


#Introduction
##DS-AD stands for "Down syndrome-associated Alzheimer's disease." This term refers to the development of Alzheimer's disease (AD) in individuals with Down syndrome (DS). People with Down syndrome have a higher risk of developing Alzheimer's disease due to the presence of an extra copy of chromosome 21, which carries the gene for amyloid precursor protein (APP). The overexpression of this gene leads to an increased production of amyloid-beta, a key protein involved in the formation of amyloid plaques, a hallmark of Alzheimer's disease. As a result, many individuals with Down syndrome develop Alzheimer's disease at an earlier age compared to the general population. The study of DS-AD is important for understanding both the unique aspects of Alzheimer's disease in Down syndrome and the general mechanisms of Alzheimer's disease in the broader population. This research can provide insights into potential therapeutic targets and preventive strategies.

##Overview of analysis
##The goal of the study "Bulk RNA-seq of sporadic Alzheimer's disease and Down syndrome with Alzheimer's disease brain tissue for prefrontal cortex and cerebellum" is to identify and analyze differential gene expression patterns in the brain tissues of individuals with sporadic Alzheimer's disease (AD) and Down syndrome-associated Alzheimer's disease (DS-AD). By focusing on the prefrontal cortex and cerebellum, the research aims to uncover the molecular mechanisms and pathways involved in the progression of Alzheimer's in these specific brain regions, which are critical for understanding the disease's impact on cognitive functions. This investigation seeks to contribute to the identification of potential biomarkers and therapeutic targets that could aid in the development of effective treatments for both sporadic AD and DS-AD.

#Analysis
####Import and Data (More Explanation to be Added)
Here is a description of the variables:
  ---------------| ---------------------------------------------------------------------------
`ID`                     | Identification
`link to article`        | https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-14179?query=E-MTAB-14179 
`Organism`               | Homo Sapiens 
`M/F`                    | Gender (M if Male, F if Female)
`Age`                    | 49-99years 
`Disease, syndrome`      | Down syndrome-associated Alzheimer's disease
`Number of participants` | 42
`Number of samples`      | 84 
`organism part`          | Cerebellum,Prefrontal Cortex 
`Extract Name`           | CTL 1 Cbl, CTL 1 Ctx, CTL 2 Cbl
`Material Type`          | RNA
`Assay Name`             | Sequencing assay
`Technology Type`        | RNA-Seq 
`Derived Array Data File`| DSAD-ApoE125_readcounts.txt 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
# Loading necessary libraries
if(!require(plyr)) install.packages("plyr")
if(!require(corrplot)) install.packages("corrplot")
if(!require(caret)) install.packages("caret")
if(!require(caretEnsemble)) install.packages("caretEnsemble")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(dplyr)) install.packages("dplyr")
if(!require(randomForest)) install.packages("randomForest")
if(!require(xgboost)) install.packages("xgboost")

# Loading libraries
library(plyr)
library(corrplot)
library(caret)
library(caretEnsemble)
library(ggplot2)
library(dplyr)
library(randomForest)
library(xgboost)

getwd()
```

```{r}
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)

metadata <- read.delim("E-MTAB-14179.sdrf.txt", sep = "\t", fill = TRUE)

head(metadata)
str(metadata)
summary(metadata)

# Filter metadata for samples with "Alzheimer's disease" and age greater than 70
filtered_metadata <- subset(metadata, Factor.Value.disease. == "Alzheimer's disease" & Characteristics.age. > 70)

head(filtered_metadata)
summary(filtered_metadata)

# Load the expression data
expression_data <- read.delim("DSAD-ApoE125_readcounts_fixed.txt", sep = "\t", header = TRUE)


head(expression_data)
str(expression_data)




```


```{r}
library(readr)
library(dplyr)

data <- read_delim("DSAD-ApoE125_readcounts_fixed.txt", delim="\t", skip=1)

# Inspect the structure and column names
str(data)
colnames(data)

# Since 'Chr', 'Start', 'End', and 'Length' do not exist in the dataset, 

data <- data %>%
  mutate(
    Chr = as.numeric(gsub(";", "", ENSG00000160072)),  # Replace with actual logic for 'Chr'
    Start = as.numeric(1),  # Placeholder value, replace with actual logic for 'Start'
    End = as.numeric(1000001),  # Placeholder value, replace with actual logic for 'End'
    Length = End - Start  # Assuming Length is calculated as 'End - Start'
  )

# Filter data for chromosome 1 and position greater than 1,000,000
filtered_data <- data %>%
  filter(Chr == 1 & Start > 1000000)

# Inspect the filtered data
head(filtered_data)

# Filter data for rows where 'Length' is greater than 1000
filtered_data <- data %>%
  filter(Length > 1000)

missing_values_summary <- filtered_data %>%
  summarise(across(everything(), ~sum(is.na(.))))

print(missing_values_summary)

write_delim(filtered_data, "filtered_data.txt", delim = "\t")



```

```{r}

dim(expression_data)
dim(filtered_metadata)

```

```{r}
expression_data <- read.delim("DSAD-ApoE125_readcounts_fixed.txt", sep = "\t", header = TRUE)

ncol(expression_data)


```
```{r}
install.packages("here")


```

```{r}
sdrf_data <- read.table("E-MTAB-14179.sdrf.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

library(dplyr)
library(here)

sdrf_data <- read.table(here("E-MTAB-14179.sdrf.txt"), header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

filtered_sdrf_data <- sdrf_data %>%
  filter(grepl("1.fq.gz$", Scan.Name))

head(filtered_sdrf_data)


```

```{r}
expression_data <- read.table("DSAD-ApoE125_readcounts_fixed.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

colnames(expression_data)


```

```{r}
expression_samples <- colnames(expression_data)[-1]  # Исключаем первую колонку


```

```{r}
metadata_samples <- filtered_sdrf_data$Scan.Name
common_samples <- intersect(expression_samples, metadata_samples)
print(common_samples)  # Посмотрим, какие образцы совпадают



```

```{r}
# Normalization of sample names from count matrix (remove prefixes, symbols, convert to lowercase)
cleaned_expression_samples <- tolower(trimws(gsub("_.*", "", expression_samples)))

cleaned_metadata_samples <- tolower(trimws(gsub("_.*", "", metadata_samples)))

common_samples <- intersect(cleaned_expression_samples, cleaned_metadata_samples)

# Print common samples
print(common_samples)

```

```{r}
# Original names from count matrix
expression_samples <- colnames(expression_data)[-1]  # Exclude the "Geneid" column


original_sample_names <- expression_samples[match(common_samples, tolower(trimws(gsub("_.*", "", expression_samples))))]

# Add the Geneid column to the filtered data
expression_data_filtered <- expression_data[, c("Geneid", original_sample_names)]

print(dim(expression_data_filtered))
head(expression_data_filtered)


```
```{r}
metadata_filtered <- filtered_sdrf_data[filtered_sdrf_data$Scan.Name %in% common_samples, ]

print(dim(metadata_filtered))
head(metadata_filtered)

```

```{r}
cleaned_metadata_samples <- tolower(trimws(gsub("_.*", "", filtered_sdrf_data$Scan.Name)))

metadata_filtered <- filtered_sdrf_data[cleaned_metadata_samples %in% common_samples, ]

print(dim(metadata_filtered))
head(metadata_filtered)

```

```{r}
print(colnames(expression_data))  # Проверьте оригинальные имена образцов


```

```{r}

# Remove suffixes (for example, "_2") from sample names count matrix
normalized_expression_samples <- gsub("_.*", "", colnames(expression_data)[-1])

print(normalized_expression_samples)

```

```{r}

# Install DESeq2 if not already installed
if(!require(DESeq2)) {
    install.packages("BiocManager")
    BiocManager::install("DESeq2")
}

library(DESeq2)
library(dplyr)




```

```{r}

normalized_expression_samples <- tolower(trimws(colnames(expression_data)))
normalized_metadata_samples <- tolower(trimws(gsub("_.*", "", filtered_metadata$Scan.Name)))

common_samples <- intersect(normalized_expression_samples, normalized_metadata_samples)

common_samples_upper <- toupper(common_samples)

common_samples_in_expression <- common_samples_upper[common_samples_upper %in% colnames(expression_data)]

if (length(common_samples_in_expression) > 0) {
  # Filter data by matching samples
    expression_data_filtered <- expression_data[, c("Geneid", common_samples_in_expression)]
    filtered_metadata_filtered <- filtered_metadata[normalized_metadata_samples %in% tolower(common_samples_in_expression), ]
    
    # Set gene identifiers using the "Geneid" column
    if ("Geneid" %in% colnames(expression_data_filtered)) {
        rownames(expression_data_filtered) <- expression_data_filtered$Geneid
        expression_data_filtered <- expression_data_filtered[, -1]  # Удаляем колонку с генами
    } else {
        print("Gene ID column 'Geneid' not found")
    }

    print(dim(expression_data_filtered))
    print(dim(filtered_metadata_filtered))
} else {
    print("No common samples found between expression_data and filtered_metadata")
}



```

```{r}
dim(expression_data)  
dim(filtered_metadata)  
```

```{r}

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!require(DESeq2)) {
  BiocManager::install("DESeq2")
}
if (!require(biomaRt)) {
  BiocManager::install("biomaRt")
}
if (!require(dplyr)) {
  install.packages("dplyr")
}
if (!require(tibble)) {
  install.packages("tibble")
}

library(DESeq2)
library(dplyr)
library(biomaRt)
library(tibble)

meta <- read.delim("E-MTAB-14179.sdrf.txt", sep="\t") %>%
  filter(grepl("1.fq.gz$", Scan.Name)) %>%
  mutate(sample_id = gsub("_1.fq.gz$", "", Scan.Name))

rownames(meta) <- meta$sample_id

meta <- dplyr::select(meta, starts_with("Characteristics"))

head(meta)



```

```{r}

# Load expression data
counts <- read.table("DSAD-ApoE125_readcounts_fixed.txt", header = TRUE, sep = "\t", row.names = 1)

# Check that samples match between expression data and metadata
common_samples <- intersect(rownames(meta), colnames(counts))

counts_filtered <- counts[, common_samples]
meta_filtered <- meta[common_samples, ]

all(rownames(meta_filtered) == colnames(counts_filtered))

head(counts_filtered)
head(meta_filtered)


```

```{r}

pacman::p_load("here", "tidyverse", "DESeq2", "ashr", "janitor", "biomaRt")

# Loading and clearing metadata
meta <- read.delim(here("E-MTAB-14179.sdrf.txt"), header=TRUE, sep="\t") %>%
  filter(grepl("1.fq.gz$", Scan.Name)) %>% 
  mutate(sample_id = gsub("_1.fq.gz$", "", Scan.Name)) %>% 
  column_to_rownames(var = "sample_id") %>% 
  dplyr::select(starts_with("Characteristics"))

# Clean up metadata column names
colnames(meta) <- gsub("^Characteristics\\.|\\.$", "", colnames(meta))

# Re-encoding the values in the disease variable for convenience
meta <- meta %>%
  mutate(disease = recode(disease,
                          "Down syndrome with Alzheimer's disease" = "down_alzh",
                          "Alzheimer's disease" = "alzh",
                          "normal" = "control")) %>% 
  mutate(organism_part = recode(organism_part,
                          "prefrontal cortex" = "prefrontal_cortex"))

unique(meta$disease)


```

```{r}

# Loading metadata
meta <- read.delim("E-MTAB-14179.sdrf.txt", sep="\t") %>%
  filter(grepl("1.fq.gz$", Scan.Name)) %>%
  mutate(sample_id = gsub("_1.fq.gz$", "", Scan.Name)) %>%
  column_to_rownames(var = "sample_id") %>%
  dplyr::select(starts_with("Characteristics"))

# Clean up metadata column names
colnames(meta) <- gsub("^Characteristics\\.|\\.$", "", colnames(meta))

meta <- meta %>%
  mutate(disease = recode(disease,
                          "Down syndrome with Alzheimer's disease" = "down_alzh",
                          "Alzheimer's disease" = "alzh",
                          "normal" = "control")) %>%
  mutate(organism_part = recode(organism_part,
                                "prefrontal cortex" = "prefrontal_cortex"))

# Load expression data
counts <- read.table("DSAD-ApoE125_readcounts_fixed.txt", header = TRUE, sep = "\t", row.names = 1)

# Testing samples for consistency
common_samples <- intersect(rownames(meta), colnames(counts))
counts_filtered <- counts[, common_samples]
meta_filtered <- meta[common_samples, ]

all(rownames(meta_filtered) == colnames(counts_filtered))



```


```{r}
# DESeq2 
dds <- DESeqDataSetFromMatrix(countData = counts_filtered, 
                              colData = meta_filtered, 
                              design = ~ disease + organism_part + organism_part:disease)




```

```{r}
dds <- DESeq(dds)


res <- results(dds, contrast = c("disease", "down_alzh", "control"))

# Print the results
res_filtered <- res %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>%
  arrange(log2FoldChange)

# Output of significant genes
head(res_filtered)

```

```{r}

plotMA(res, ylim = c(-2, 2))

# Volcano plot
res$Significant <- ifelse(res$padj < 0.05 & abs(res$log2FoldChange) > 1, "Yes", "No")
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log10(pvalue), color = Significant)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  labs(title = "Volcano plot", x = "Log2 fold change", y = "-Log10 p-value")


```
#X-axis: Displays the mean of normalized counts, representing the average gene expression levels. It shows how strongly genes are expressed on average.

#Y-axis: Displays the log fold change, indicating the difference in expression levels between two conditions (e.g., control and disease).

#Results: Points that are far from the horizontal axis (log fold change = 0) indicate genes with significant changes in expression.

#If a point is above zero, it means the gene is expressed higher in one of the conditions.If a point is below zero, the gene is expressed lower.Genes with small changes: Most points are grouped around the Y-axis = 0, indicating that most genes have minimal changes in expression between conditions.

#Point density: A high density of points near zero on the Y-axis and in the center of the X-axis suggests that most genes have moderate expression levels and show no significant changes between conditions.

#Highlighted genes (blue points): Blue points may indicate genes that deviate from the overall trend and may have significant changes in expression, requiring further analysis.

#Conclusion:Most genes show either minimal changes in expression or low expression levels. Genes that stand out on this graph (with significant log fold change) may be biologically significant and may require further investigation.
```{r}
# Save results to CSV file
write.csv(res_filtered, "deseq2_results_filtered.csv", row.names = TRUE)
```